\documentclass[a4paper, twocolumn]{article}

% you can switch between these two (and more) styles by commenting one out (use percentage)
\usepackage[backend=biber]{biblatex}
%\usepackage[backend=biber, style=authoryear-icomp]{biblatex}
\addbibresource{./refs.bib}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{color}
\definecolor{lightgray}{gray}{0.9}
% code listing: https://tex.stackexchange.com/questions/19004/how-to-format-an-inline-source-code
\lstset{
    showstringspaces=false,
    basicstyle=\ttfamily,
    keywordstyle={blue},
    commentstyle=\color[gray]{0.6}
    stringstyle=\color[RGB]{255, 150, 75}
}
\newcommand{\inlinecode}[2]{\colorbox{lightgray}{\lstinline[language=#1]$#2$}}
\author{Jeffrey Roed, Rasmus Kibshede, Joachim Richter}
\title{Genetic algorithms snake game}

\begin{document}
\twocolumn[
    \begin{@twocolumnfalse}
        \maketitle
        \begin{abstract}
            TBA
        \end{abstract}
    \end{@twocolumnfalse}
    \vspace{1cm}
]

\section {Introduction\label{sec:Introduction}}
 
The primary objective of this study is to demonstrate how genetic algorithms (GAs) can be utilized to develop an agent that plays the Snake game. The Snake game, a classic arcade game, involves navigating a snake to collect food items while avoiding collisions with walls and the snake's own body. The GA is used to optimize the snake's behavior over successive generations, improving its ability to survive and collect food.
\subsection{Research question\label{sec:Research Question}}
How can we develop a genetic algorithm that not only matches but exceeds our group's average score in the game of Snake by leveraging evolutionary strategies, including effective initialization, precise fitness evaluation, robust selection, and strategic use of crossover and mutation?

\section{Method\label{sec:Method}}
\subsection{Selection\label{sec:Selection}}

The selection methods is responsible for choosing parent chromosomes that will generate the next generation of individuals. Below are some thoughts we have made in choosing the right selection methods for our Snake game.

\begin{itemize}
\item \textbf{Roulette Wheel Selection.}
Individuals are chosen with a probability proportional to their fitness. This maintains genetic diversity and prevents premature convergence \cite{goldberg1991comparative}.
        
Using this method early in the training process is advantageous when the fitness landscape is smooth and diverse strategies need to be explored.

\item \textbf{Tournament Selection.}
Randomly selected individuals compete, and the winner is chosen for breeding.

Will be useful when the algorithm progresses and a more refined selection is needed to accelerate the optimization process \cite{goldberg1991comparative}.

\item \textbf{Rank Selection.}
Individuals are ranked based on fitness values, and selection probabilities are based on ranks. This reduces the risk of premature convergence.

Will be useful throughout the evolutionary process by providing steady selection pressure and maintaining diversity \cite{srinvas1994adaptive}.

\end{itemize}

To determine the most effective selection method for our GA, we propose to evaluate each approach by conducting multiple runs of the Snake game simulation. The performance will be assessed based on the average score achieved by the evolved agents over a fixed number of generations. This will help us determine the most suitable strategy for our Snake game agent. Adaptive approaches to selection, as discussed by Srinivas and Patnaik (1994), may also be considered to enhance the overall performance and robustness of the algorithm \cite{srinvas1994adaptive}.

\subsection{Fitness\label{sec:Fitness}}
The fitness function evaluates how well each individual in a population solves the given problem, assigning a numerical score based on their performance. This score guides the selection process for reproduction, favoring higher-scoring individuals to pass their traits to the next generation. By driving the evolutionary progress and maintaining diversity, the fitness function helps the algorithm converge toward optimal or near-optimal solutions

\subsection{Crossover\label{sec:Crossover}}
The crossover methods are used in GAs to combine the genetic information of two parent chromosomes in order to generate new offspring. Below are some thoughts we have made in choosing the right crossover methods for our Snake game.

\begin{itemize}
    \item \textbf{Single point Crossover}
A random crossover point is selected on the parent chromosomes. All of the genetic information is copied from one parent, and all the genetic information after this point is copied from the other parent. This is a  simple method that keeps the genetic information of both parents making is useful for upholding diversity in the population. 

This will be beneficial in the early stages of training, when we need to explore diverse strategies. 
\end{itemize}


\begin{itemize}
    \item \textbf{Two-Point Crossover}
Selects two random crossover point on the parent chromosomes. The genetic information is interchanged between the parents to create two new offspring. This method provides a higher level of mixing compared to single-point crossover, which can lead to more varied breeding. 

This will be useful in later stages of training, where we need to fine tune the algorithm strategies by combining the most desirable features of parents. 
\end{itemize}

As highlighted by Henrik Strøm \cite{strom2024travelling}, the biggest challenge in implementing crossover for problems such as Traveling Salesman Problem (TSP), is building an effective crossover mechanism. TSP is about finding the shortest possible route to visit a number of cities only once,and afterwards return to the starting city.
Similar considerations apply to our Snake game GA, where the crossover method must effectively combine parent strategies without breaking the underlying structure of the Nerual Network (NN).

In order to address these challenges, we will also explore crossover methods designed for TSP, such as; Order Crossover, and Partially Mapped Crossover, which maintains validity of the offspring. We will during our research experiment with different techniques in order to find the most optimal crossover method.  

\subsection{Mutation\label{sec:Mutation}}
 
The mutation function aims to modify the genome of the offspring. These modifications serve to maintain and introduce diversity in the genetic population, which is crucial for the overall success of the genetic algorithm, and prevents premature convergence.
\begin{quote}
\emph{
    "The role of mutation in GAs has been that of restoring lost or unexplored genetic material into the population to prevent the premature convergence of the GA to suboptimal solutions."
    }
    \cite{srinvas1994adaptive}
\end{quote}


\subsection{Mutation V.2 Jeff\label{sec:Mutation}}

Mutation is a critical component of genetic algorithms (GAs), serving to introduce genetic diversity into the population and prevent premature convergence to suboptimal solutions. By altering the genetic material of offspring, mutation helps maintain a healthy level of variability, ensuring the exploration of a broader solution space.

David E. Goldberg, in his foundational text \textit{Genetic Algorithms in Search, Optimization, and Machine Learning}, emphasizes the importance of mutation in maintaining genetic diversity and aiding the convergence of the algorithm to optimal solutions. He notes that mutation plays a crucial role in restoring lost or unexplored genetic material, thus enhancing the robustness of GAs \cite{goldberg1989genetic}.

Robert E. Marks, in \textit{Playing Games with Genetic Algorithms}, highlights the application of GAs in game strategies, noting that mutation is essential for evolving strategies that can adapt to dynamic environments and unforeseen challenges \cite{marks2001playing}.

\textbf{Types of Mutation:}
\begin{itemize}
    \item \textbf{Bit Flip Mutation:} Common in binary-encoded GAs, this method involves flipping one or more bits in a chromosome. Simple yet effective for maintaining genetic diversity. Suitable for early stages of evolution where broad exploration is needed.
    \item \textbf{Swap Mutation:} Involves swapping two genes within a chromosome. Particularly useful for problems like the Traveling Salesman Problem (TSP) and similar permutation-based problems.
    \item \textbf{Scramble Mutation:} Selects a subset of genes and randomly scrambles them. Maintains the overall genetic material but alters the order, fostering diversity.
    \item \textbf{Gaussian Mutation:} Applied to real-valued GAs, where a small Gaussian noise is added to the genes. Effective for fine-tuning solutions in continuous optimization problems.
\end{itemize}

\begin{quote}
\emph{"The role of mutation in GAs has been that of restoring lost or unexplored genetic material into the population to prevent the premature convergence of the GA to suboptimal solutions."} \cite{srinvas1994adaptive}
\end{quote}

\textbf{Implementing Mutation in the Snake Game GA:} To enhance the performance of our Snake game agent, we propose to experiment with various mutation techniques, monitoring their impact on the agent's ability to adapt and improve over successive generations. By fine-tuning the mutation rate and method, we aim to balance exploration and exploitation, leading to a more robust and high-performing Snake game strategy.




\section{Analysis\label{sec:Analysis}}
\subsection{Selection\label{sec:Selection}}
\subsection{Fitness\label{sec:Fitness}}
\subsection{Crossover\label{sec:Crossover}}
\subsection{Mutation\label{sec:Mutation}}
single point mutation

\section{Findings\label{sec:Findings}}
\subsection{Selection Enhancement\label{sec:Selection Enhancement}}
\subsection{Crossover Enhancement\label{sec:Crossover Enhancement}}
\subsection{Mutation Enhancement\label{sec:Mutation Enhancement}}
\printbibliography
\end{document}
